openarchaeoCollaboration

### 2023-03-31 analysis/06-project_activity.qmd
Updating the text of analysis and interpretation.
Re-wording for more active voice, adding some interpretation (specifically to section 07), and updates the notes for further work to be done.

In this section we assess projects' pulse, or the amount of activity pertaining to a repository. In the following figure, each vertical line represents the lifespan of a repository, and the points on the line are individual commits. The graph is sorted by the date of the most recent commit for each repository, so the ones on the top became inactive earlier and the ones on the bottom represent ongoing work.

{r plot-lifespan}

The span of time between the first and latest commits is visualized more clearly in summary here. Notably, many projects are active for only a few short period of time: about 15% less than a month, 25% less than three months, and 40% less than a year. Projects with a short span of activity tend to have very high commit frequency, whereas there is much more variety in long-term projects. Among projects with a greater lifespan, those with more total commits have much more frequent activity than those with fewer total commits, which represent the majority of long-lived projects. The overall negative trend (seen in the top-right graph) can be explained in terms of natural processes, i.e. the only way to fit a high number of commits within the same span as a project with a low number of commits is to commit more frequently. But bearing in mind both axes use a logarithmic scale, the upper left portion of the graph does appear to suggest that short-lived projects are characterised by a spree of rapid committing; while longer-lived projects have a more diverse range of workrates. This latter pattern in turn appears to be explained by the number of contributors: projects that sustain a high commit rate over a long period of time tend to rely on a large number of contributors, as per the bottom-right graph.

{r lifespan-stats}

{r plot-lifespan-hist}

{r plot-lifespan-vs-rate}

Perhaps unfortunately then, we also find that nearly 60% of projects have only a single contributor, and projects with more than a dozen are very rare indeed \[\<3%\]. A "contribution", in this case, is a commit, issue, comment applied by a user to a repository.

@@

Furthermore, when we break down the contributions made by all contributors in multi-contributor projects, we also see that work tends to be distributed very unevenly. Of multi-contributor repositories, 92% have a third of contributions made by a single person; 65% have half of all contributions made by a single person; and 36% have three quarters of all contributions made a single person.

@@

{r plot-issues-hist}

{r plot-comments-hist}

>>>>

But how is this activity distributed across repositories? In this section we assess projects' pulse, or the amount of activity pertaining to a repository.

In the following figure, each vertical line represents a repository's lifespan, and the points on each line are individual commits. The graph is sorted by the date of the most recent commit for each repository, so the ones on the top became inactive earlier and the ones on the bottom represent ongoing work.

{r plot-lifespan}
{r lifespan-stats}

The span of time between the first and latest commits is visualized more clearly in the following charts. Notably, many projects are active for only a short period of time: about 15% less than a month, 25% less than three months, and 40% less than a year. Projects with a short span of activity tend to have very high commit frequency, whereas there is much more variety in long-term projects. Among projects with a greater lifespan, those with more total commits have much more frequent activity than those with fewer total commits, which represent the majority of long-lived projects. The overall negative trend (seen in the top-right graph) can be explained in terms of natural processes, i.e. the only way to fit a high number of commits within the same span as a project with a low number of commits is to commit more frequently. But bearing in mind both axes use a logarithmic scale, the upper left portion of the graph does appear to suggest that short-lived projects are characterised by a spree of rapid committing; while longer-lived projects have a more diverse range of workrates. This latter pattern in turn appears to be explained by the number of contributors: projects that sustain a high commit rate over a long period of time tend to rely on a large number of contributors, as per the bottom-right graph.

**NOTE:** Combine these figures using patchwork.
{r plot-lifespan-hist}
{r plot-lifespan-vs-rate}

Perhaps unfortunately then, we also find that nearly 60% of projects have only a single contributor, and projects with more than a dozen are very rare indeed \[\<3%\]. A "contribution", in this case, is a commit, issue, or comment applied by a user to a repository.

@@

Furthermore, when we break down the contributions made by all contributors in multi-contributor projects, we also see that work tends to be distributed very unevenly. Of multi-contributor repositories, 92% have a third of contributions made by a single person, 65% have half of all contributions made by a single person, and 36% have three quarters of all contributions made a single person.

**NOTE:** Re-calculate these numbers to ensure they are up to date. Translate the first point to be about how many repos have a quarter of contributions by a single person, so that it corresponds with the vertical lines on the graphic, which are now divided into quarters rather than thirds.

@@

**NOTE:** Combine these figures using patchwork.
{r plot-issues-hist}
{r plot-comments-hist}

### 2023-03-31 analysis/07-network.qmd
Updating the text of analysis and interpretation.
Re-wording for more active voice, adding some interpretation (specifically to section 07), and updates the notes for further work to be done.

We constructed and analyzed networks representing users' contributions to GitHub repositories in order to examine the emergence of collaborative tendencies. We sought to identify clusters of repositories that share common sets of contributors, and the different kinds of contributions that draw the network together.

@@

We started by constructing a graph connecting users to the repositories that they contributed to, accounting for commits, issues, and comments as distinct kinds of relations. We then extracted two one-mode networks from this, one connecting repositories by common users, and the other connecting users by common repositories.

@@

### repo-repo
We applied the edge-betweenness community detection method to identify clusters of repositories that share common sets of contributors of all kinds. Aside from isolate notes, which represent repositories with only a single contributor, we detected 21 clusters. While many of these clusters are interconnected, some discrete components containing between 2-7 repositories appear as distinct from a primary core.

Are the clusters associated with particular topics or tags? Are there some sorts of associations between clusters that break down from a parent in the community detection dendrogram? How does all of this differ when we differentiate commits from issues and comments?

{r plot-graph-repo-repo-edge-betweenness}

{r plot-graph-repo-repo-edge-betweenness-dendrogram}

* Identify the repos/organisations that are the focus of the most intense collaboration?
  * Are they thematically or topically linked?
  * Do they map onto real-world research centres?
* Is the network a new thing, or just a translation of existing offline networks (e.g. fields, institutions) onto GitHub?
  * cf. is open source really open?

### user-user

We assembled graphs linking users based on common contributions to the same repositories. We sought to identify whether certain users, who contribute in certain distinct ways, play different roles in overall network.

We found that the people with the highest betweenness values are those who primarily produce computational archaeology code as a job.

We applied the same betweenness centrality algorithm on user-user networks whose links are based on only commits and on only issues and comments. For the latter, people who commit less code are more represented with higher betweenness scores, but many of the people with high betweenness from the all-contributions graph also appear here. These people who appear in both of these lists have a tendency to contribute as both committers and as commenters.

{r graph-user-user}

@@

{r plot-graph-user-user}

{r plot-user-betweenness-dist}

{r plot-user-betweenness-activity}

@@

>>>>

We constructed and analyzed networks representing users' contributions to GitHub repositories to examine the emergence of collaborative tendencies. We sought to identify clusters of repositories that share common sets of contributors, and the different kinds of contributions that draw the network together.

@@

We started by constructing a graph connecting users to the repositories that they contributed to, accounting for commits, issues, and comments as distinct kinds of relations.

@@

We then extracted two one-mode networks from this, one connecting repositories by common users, and the other connecting users by common repositories.

**NOTE:** Combine these figures using patchwork.
**NOTE:** Integrate {r graph-user-user} and {r plot-graph-user-user} into a single cell, similar to how {r plot-graph-repo-repo} integrates both components.

{r plot-graph-repo-repo}
{r graph-user-user}
{r plot-graph-user-user}

## repo-repo
We applied the edge-betweenness community detection method to identify clusters of repositories that share common sets of contributors of all kinds. Aside from isolate nodes, which represent repositories with only a single contributor, we detected 21 clusters. While many of these clusters are interconnected, some discrete components containing between 2-7 repositories appear as distinct from a primary core.

{r plot-graph-repo-repo-edge-betweenness}
{r plot-graph-repo-repo-edge-betweenness-dendrogram}

This graph identifies three core clusters characterized by repositories whose contributors commit to projects other than their own. The clusters that appear along the peiphery largely correspond with the work of single individuals, and sometimes also their colleagues. Peripheral clusters that are connected to core clusters by only a few relationships represent the sole (and perhaps initial) integrations of lone developers into a broader community.^[**NOTE:** I wonder what these repositories that prompt greater integration are? They should correspond with the nodes of highest betweenness centrality among those in the core cluster.]^[**NOTE:** I also wonder how recently these connections were forged, or how they are situated within users' history of commits.]

The three core clusters also have distinct character. One is characterized by its focus on archaeogenetics, which consists of a very well established collaborative network and general reliance on data modelling and processing tools. A second cluster is mostly centred on fieldwork-oriented data collection tools, and particularly the emergence of dominant platforms that attract more attention than other independent projects scattered across the network. The third and most significant cluster includes a schmorgasborg of projects whose contributors share varied interests. The emphasis in this latter cluster is on the formation of a central software development community rather than on any specific topic of work.^[**NOTE:** How does this map onto real-world research centres or partnerships? I'm wary of getting too into the weeds on specific people or relationships, but we might highlight the fact that projects led by people who are more active in the SSLA are well represented in this cluster, as well as members of ISAA-Kiel. Note that this observation is not backed by statistics, it is just a casual remark based on me browsing the network visualization.]

**NOTE:** What happens to this interpretation when we differentiate commits from issues and comments?

## user-user

We assembled graphs linking users based on common contributions to the same repositories. We sought to identify whether certain users, who contribute in certain distinct ways, play different roles in overall network. We found that the people with the highest betweenness values are those who primarily produce computational archaeology code as their job.

**NOTE:** Create a graph visualization showing either edge-betweenness clustering or rendering nodes' relative betweenness by size.
**NOTE:** We can further investigate this point made above by comparing with some other measures of activity, like volume of commits.

We applied the same betweenness centrality algorithm on user-user networks whose links are based on only commits and on only issues and comments. For the latter, people who commit less code are more represented with higher betweenness scores, but many of the people with high betweenness from the graph representing all contributions also appear here. These people who appear in both of these lists have a tendency to contribute as both committers and as commenters.

@@

**NOTE:** I have no idea what these next two charts depict.

{r plot-user-betweenness-dist}
{r plot-user-betweenness-activity}

---

@@

---

In cleaning the data we made a few significant decisions that revealed insights into our dataset. We constructed the user-user graph and filtered out several bots. We filtered out several large non-archaeology projects that effected the network in ways that distracted us from what we wanted to say about the archaeological community. We filtered isolates. We calculated the degree and betweenness to assess whether this was a realistic representation of what we know. We wanted to figure out who are the key nodes, what they had in common.

The people who are high betweenness are people who primarily produce computational archaeology code as a job. We can investgate further by comparing with some other measures of activity, like volume of commits or something else. We also looked at the same thing, but with only commits, or without commits. For the subgraph of non-commits, it includes many people from the commits-only or commits-included graphs, but also more people who do not commit as much.

### 2023-03-31 analysis/99-conclusion.qmd
Updating the text of analysis and interpretation.
Re-wording for more active voice, adding some interpretation (specifically to section 07), and updates the notes for further work to be done.

Collaborative open source software development in archaeology has seen a rapid and sustained increase beginning around 2014 or 2015.

In terms of collaboration, our analysis shows an uneven use of git and GitHub's extended features, beyond their basic usage as a version control system and repository host.

The vast majority of repositories have 1--3 contributors, with only a few distinguished by an active and diverse developer base.

Generally speaking, we believe that this is because people do not want to step on other people's shoes by raising issues or intruding on other people's projects.

This may relate to our positions as academics who hold different values relative to the designers of open source development environments, regarding how collaboration should occur, for example, when dealing with how projects and ideas are 'owned' by individuals or communities, and how work should be iteratively improved upon.

We intend to look deeper at the social aspects of collaborative software engineering among archaeologists with a network analysis, and perhaps by leveraging more data from documentation included in each repository.

---

There is an emerging community of practice around open source in archaeology.

But:

* It is limited and fragmented
  * A small set of subfields are represented
  * Probably most people still haven't heard it?
* Collaboration is limited
  * Most work remains solo & short-lived
  * Perhaps because open source norms conflict with academic norms?
* The members of this community face structural challenges in academia

>>>>

Collaborative open source software development in archaeology has seen a rapid and sustained increase beginning around 2014 or 2015. This is marked by a variety of applications and use cases. While there is an emerging community of practice around open source in archaeology, collaboration remains limited. Most work is performed solo and is short-lived.^[**NOTE:** I wonder if we can say something about change over time, for instance, if things are getting better.]

The vast majority of repositories have 1--3 contributors, with only a few distinguished by an active and diverse developer base. Our analysis also shows an uneven use of git and GitHub's extended features, beyond their basic usage as a version control system and repository host. Generally speaking, we believe that this is because people do not want to step on other people's shoes by raising issues or intruding on other people's projects. This may relate to our positions as academics who hold different values relative to the designers of open source development environments, regarding how collaboration should occur, for example, when dealing with how projects and ideas are 'owned' by individuals or communities, and how work should be iteratively improved upon.
